---
title: "gene_identification"
author: "Erica Robertson"
date: "2024-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries}
library(dplyr)
library(tidyverse)
```

Generic commands for getting out the SNP bed files. Might be an easier way but this was simple enough...
```{generic_code}
cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E 
"SNP_1|SNP_2" > 
../new_input_files/snps/trait_model_snps.bed

bedtools sort -i trait_model_snps.bed > trait_model_snps_sorted.bed

bedtools closest -a trait_model_snps_sorted.bed -b 
/scratch/alpine/ericacnr@colostate.edu/GCRF/Reference/ZEFI_annotation/Taeniopygia_guttata_sorted.bed 
-d > trait_model_distance.bed
```

For loops to generate the actually commands used
```{r lmm_commands}
feather_traits <- c("pen_num", "pen_length", "plum_length", "plum_num", "nodes", "GCRF_feather_PC")
for(trait in feather_traits) {
  # Construct the file path
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_feather/", trait, "_outlier.txt")
  
  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the significant SNPs from the file
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    
    # Concatenate SNP names
    concatenated_snps <- paste(snp_names, collapse = "|")
    
    # Construct the command
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_lmm_snps.bed")
    
    # Print the command
    print(command)
  } else {
    # File does not exist, print a message and skip to the next trait
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}

beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_bill/", trait, "_outlier.txt")
  if (file.exists(file_path)) {
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    concatenated_snps <- paste(snp_names, collapse = "|")
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_lmm_snps.bed")
    print(command)
  } else {
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}
```
```{r bslmm_commands}
feather_traits <- c("pen_num", "pen_length", "plum_length", "plum_num", "nodes", "GCRF_feather_PC")
for(trait in feather_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_feather/", trait, "_pip01.dsv")
  if (file.exists(file_path)) {
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    concatenated_snps <- paste(snp_names, collapse = "|")
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_bslmm_snps.bed")
    print(command)
  } else {
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}

beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_beak/", trait, "_pip01.dsv")
  if (file.exists(file_path)) {
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    concatenated_snps <- paste(snp_names, collapse = "|")
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_bslmm_snps.bed")
    print(command)
  } else {
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}
```

For loops to run in the command line
```{unix_commands}
for trait in `ls *_snps.bed | rev | cut -d'_' -f2- | rev`; do echo $trait; bedtools sort -i "$trait"_snps.bed > "$trait"_snps_sorted.bed; done

for trait in `ls *_sorted.bed | rev | cut -d'_' -f3- | rev`; do echo $trait; bedtools closest -a "$trait"_snps_sorted.bed -b /scratch/alpine/ericacnr@colostate.edu/GCRF/Reference/ZEFI_annotation/Taeniopygia_guttata_sorted.bed -d > "$trait"_distance.bed; done
```

# FEATHER GENES
```{r feather_filtering}
feather_traits <- c("pen_num", "plum_num", "pen_length", "plum_length", "nodes", "GCRF_feather_PC")

#LMM
for(trait in feather_traits) {
  # Construct the file path
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_lmm_distance.bed")
  
  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the table from the file
    df <- read.table(file_path, sep = "\t")
    
    # Filter for genes
    trait_genes <- df %>% filter(V12 == "gene")
    
    threshold <- 100000
    
    # Filter for SNPs within 100000kb of gene start or stop, or inside the gene
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    
    # Separate the columns
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    
    # Add column names
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    
    # Combine data frames
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    
    # Extract gene names
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    
    # Write gene names to file
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_lmm_100kb.txt"), row.names=F, quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}

#BSLMM
for(trait in feather_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_bslmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_bslmm_100kb.txt"), quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}

```

#BEAK GENES
```{r beak_filtering}
beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")

#LMM
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_lmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_lmm_100kb.txt"), quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}

#BSLMM
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_bslmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_bslmm_100kb.txt"), quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}

beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")
trait <- beak_traits[1]
file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_bslmm_distance.bed")
  
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_bslmm_100kb.txt"), quote = FALSE, sep = "\t")
```


### Condensing into one tidy dataframes
```{r}
library(dplyr)
# List all files matching the pattern
bslmm_gene_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", pattern = "_genes_bslmm_100kb.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
bslmm_data <- lapply(bslmm_gene_files, function(file) {
  # Extract trait name from file name
  trait <- gsub("_genes_bslmm_100kb\\.txt", "", basename(file))
  
  # Read file
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  
  # Check number of rows
  if (nrow(data) < 2) {
    cat("Skipping file", file, "because it has less than 2 rows of data.\n")
    return(NULL)  # Skip processing and return NULL
  }
  
  # Add trait column
  data$trait <- trait
  return(data)
})

# Combine all data frames into a single large data frame
bslmm_combined <- bind_rows(bslmm_data)
write_csv(bslmm_combined, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/bslmm_genes.csv")

###########################################
lmm_gene_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", pattern = "_genes_lmm_100kb.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
lmm_data <- lapply(lmm_gene_files, function(file) {
  trait <- gsub("_genes_lmm_100kb\\.txt", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  if (nrow(data) < 2) {
    cat("Skipping file", file, "because it has less than 2 rows of data.\n")
    return(NULL)  # Skip processing and return NULL
  }
  data$trait <- trait
  return(data)
})
lmm_combined <- bind_rows(lmm_data)
write_csv(lmm_combined, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/lmm_genes.csv")
```

## Big Master Sheet...

```{r}
bslmm_beak_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_beak/", pattern = "_pip01.dsv", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
bslmm_beak <- lapply(bslmm_beak_files, function(file) {
  trait <- gsub("_pip01\\.dsv", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(bslmm_beak)
bslmm_beak_all <- bind_rows(bslmm_beak)
bslmm_beak_all$model <- "bslmm"
bslmm_beak_tidy <- bslmm_beak_all[,c(9,8,2)]
View(bslmm_beak_tidy)

bslmm_feather_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_feather/", pattern = "_pip01.dsv", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
bslmm_feather <- lapply(bslmm_feather_files, function(file) {
  trait <- gsub("_pip01\\.dsv", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(bslmm_feather)
bslmm_feather_all <- bind_rows(bslmm_feather)
bslmm_feather_all$model <- "bslmm"
bslmm_feather_tidy <- bslmm_feather_all[,c(9,8,2)]
View(bslmm_feather_tidy)

bslmm_all <- rbind(bslmm_beak_tidy, bslmm_feather_tidy)

write_csv(bslmm_all, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_snps.csv")
```

```{r}
lmm_beak_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_bill/", pattern = "_outlier.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
lmm_beak <- lapply(lmm_beak_files, function(file) {
  trait <- gsub("_outlier\\.txt", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(lmm_beak)
lmm_beak_all <- bind_rows(lmm_beak)
lmm_beak_all$model <- "lmm"
lmm_beak_tidy <- lmm_beak_all[,c(15,14,2)]
View(lmm_beak_tidy)

lmm_feather_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_feather/", pattern = "_outlier.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
lmm_feather <- lapply(lmm_feather_files, function(file) {
  trait <- gsub("_outlier\\.txt", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(lmm_feather)
lmm_feather_all <- bind_rows(lmm_feather)
lmm_feather_all$model <- "lmm"
lmm_feather_tidy <- lmm_feather_all[,c(15,14,2)]
View(lmm_feather_tidy)

lmm_all <- rbind(lmm_beak_tidy, lmm_feather_tidy)

write_csv(lmm_all, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_snps.csv")
```

```{r}
lmm_all <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_snps.csv")
bslmm_all <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_snps.csv")

all_snps <- rbind(bslmm_all, lmm_all)

lmm_genes <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/lmm_genes.csv")
bslmm_genes <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/bslmm_genes.csv")

all_genes <- rbind(bslmm_genes, lmm_genes)

master_snps <- merge(all_snps, all_genes[, c("rs", "gene_name", "gene_id")], by = "rs", all.x = TRUE)
write.csv(master_snps, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/results_master.csv", row.names=F)
```




# POPULATION LEVEL ANALYSIS
```{r lmm_commands}
all_traits <- c("pen_num", "pen_length", "plum_length", "plum_num", "nodes", "feather_PC", "beak_depth", "beak_width", "culmen_end_length", "nare_length", "bill_PC")
for(trait in all_traits) {
  # Construct the file path
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/lmm_poplevel/", trait, "_fam1_outlier.txt")
  
  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the significant SNPs from the file
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    
    # Concatenate SNP names
    concatenated_snps <- paste(snp_names, collapse = "|")
    
    # Construct the command
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_fam1_lmm_snps.bed")
    
    # Print the command
    print(command)
  } else {
    # File does not exist, print a message and skip to the next trait
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}

for(trait in all_traits) {
  # Construct the file path
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/lmm_poplevel/", trait, "_fam2_outlier.txt")
  
  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the significant SNPs from the file
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    
    # Concatenate SNP names
    concatenated_snps <- paste(snp_names, collapse = "|")
    
    # Construct the command
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_fam2_lmm_snps.bed")
    
    # Print the command
    print(command)
  } else {
    # File does not exist, print a message and skip to the next trait
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}
```

```{r bslmm_commands}
feather_traits <- c("pen_num", "pen_length", "plum_length", "plum_num", "nodes", "GCRF_feather_PC")
for(trait in feather_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_feather/", trait, "_pip01.dsv")
  if (file.exists(file_path)) {
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    concatenated_snps <- paste(snp_names, collapse = "|")
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_bslmm_snps.bed")
    print(command)
  } else {
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}

beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_beak/", trait, "_pip01.dsv")
  if (file.exists(file_path)) {
    sig_snps <- read.table(file_path, header = TRUE, sep = "\t") 
    snp_names <- sig_snps[, 2]
    concatenated_snps <- paste(snp_names, collapse = "|")
    command <- paste0("cat GCRF.SNP.imputed_gt.ZFchr.sorted.bed | grep -E '", concatenated_snps, "' > ../new_input_files/snps/", trait, "_bslmm_snps.bed")
    print(command)
  } else {
    cat("File", file_path, "does not exist. Skipping trait:", trait, "\n")
  }
}
```

For loops to run in the command line
```{unix_commands}
for trait in `ls *_fam*_lmm_snps.bed | rev | cut -d'_' -f3- | rev`; do echo $trait; bedtools sort -i "$trait"_lmm_snps.bed > "$trait"_lmm_snps_sorted.bed; done

for trait in `ls *_fam*_lmm_snps.bed | rev | cut -d'_' -f3- | rev`; do echo $trait; bedtools closest -a "$trait"_lmm_snps_sorted.bed -b /scratch/alpine/ericacnr@colostate.edu/GCRF/Reference/ZEFI_annotation/Taeniopygia_guttata_sorted.bed -d > "$trait"_lmm_distance.bed; done
```

# all genes
```{r feather_filtering}
all_traits <- c("pen_num", "pen_length", "plum_length", "plum_num", "nodes", "feather_PC", "beak_depth", "beak_width", "culmen_end_length", "nare_length", "bill_PC")

#LMM
for(trait in all_traits) {
  # Construct the file path
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam1_lmm_distance.bed")
  
  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the table from the file
    df <- read.table(file_path, sep = "\t")
    
    # Filter for genes
    trait_genes <- df %>% filter(V12 == "gene")
    
    threshold <- 100000
    
    # Filter for SNPs within 100000kb of gene start or stop, or inside the gene
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    
    # Separate the columns
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    
    # Add column names
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    
    # Combine data frames
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    
    # Extract gene names
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    
    # Write gene names to file
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam1_genes_lmm_100kb.txt"), row.names=F, quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}
for(trait in all_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam2_lmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam2_genes_lmm_100kb.txt"), row.names=F, quote = FALSE, sep = "\t")
  } else {
    cat("File", file_path, "does not exist.\n")
  }
}

#BSLMM
for(trait in all_traits) {
  # Construct the file path
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam1_bslmm_distance.bed")
  
  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the table from the file
    df <- read.table(file_path, sep = "\t")
    
    # Filter for genes
    trait_genes <- df %>% filter(V12 == "gene")
    
    threshold <- 100000
    
    # Filter for SNPs within 100000kb of gene start or stop, or inside the gene
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    
    # Separate the columns
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    
    # Add column names
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    
    # Combine data frames
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    
    # Extract gene names
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    
    # Write gene names to file
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam1_genes_bslmm_100kb.txt"), row.names=F, quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}
for(trait in all_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam2_bslmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/analysis_output_poplevel/genes/", trait, "_fam2_genes_bslmm_100kb.txt"), row.names=F, quote = FALSE, sep = "\t")
  } else {
    cat("File", file_path, "does not exist.\n")
  }
}

```

#BEAK GENES
```{r beak_filtering}
beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")

#LMM
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_lmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_lmm_100kb.txt"), quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}

#BSLMM
for(trait in beak_traits) {
  file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_bslmm_distance.bed")
  if (file.exists(file_path)) {
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_bslmm_100kb.txt"), quote = FALSE, sep = "\t")
  } else {
    # Print a message if the file doesn't exist
    cat("File", file_path, "does not exist.\n")
  }
}

beak_traits <- c("beak_depth", "beak_width", "culmen_end_length", "nare_length", "GCRF_bill_PC")
trait <- beak_traits[1]
file_path <- paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_bslmm_distance.bed")
  
    df <- read.table(file_path, sep = "\t")
    trait_genes <- df %>% filter(V12 == "gene")
    threshold <- 100000
    close_genes <- trait_genes %>%
      filter((V2 >= (V6 - threshold) & V2 <= (V6 + threshold)) |
             (V2 >= (V7 - threshold) & V2 <= (V7 + threshold)) |
             (V2 >= V6 & V2 <= V7))
    df <- separate(close_genes, V14, into = paste0("V14_", 1:6), sep = "; ")
    colnames(df)[grep("^V14_", colnames(df))] <- c("gene_id", "gene_version", "gene_name", "gene_source", "gene_biotype", "additional_info")
    close_genes <- data.frame(close_genes[,1:13], df, close_genes$V15)
    gene_names <- data.frame(close_genes$V4, close_genes$gene_id, close_genes$gene_name)
    colnames(gene_names) <- c("rs", "gene_id", "gene_name")
    just_ID <- separate(gene_names, gene_id, into=c(NA, "gene_id"), sep= " ")
    just_names <- separate(just_ID, gene_name, into = c(NA, "gene_name"), sep = " ")
    write.table(just_names, paste0("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", trait, "_genes_bslmm_100kb.txt"), quote = FALSE, sep = "\t")
```


### Condensing into one tidy dataframes
```{r}
library(dplyr)
# List all files matching the pattern
bslmm_gene_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", pattern = "_genes_bslmm_100kb.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
bslmm_data <- lapply(bslmm_gene_files, function(file) {
  # Extract trait name from file name
  trait <- gsub("_genes_bslmm_100kb\\.txt", "", basename(file))
  
  # Read file
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  
  # Check number of rows
  if (nrow(data) < 2) {
    cat("Skipping file", file, "because it has less than 2 rows of data.\n")
    return(NULL)  # Skip processing and return NULL
  }
  
  # Add trait column
  data$trait <- trait
  return(data)
})

# Combine all data frames into a single large data frame
bslmm_combined <- bind_rows(bslmm_data)
write_csv(bslmm_combined, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/bslmm_genes.csv")

###########################################
lmm_gene_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/", pattern = "_genes_lmm_100kb.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
lmm_data <- lapply(lmm_gene_files, function(file) {
  trait <- gsub("_genes_lmm_100kb\\.txt", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  if (nrow(data) < 2) {
    cat("Skipping file", file, "because it has less than 2 rows of data.\n")
    return(NULL)  # Skip processing and return NULL
  }
  data$trait <- trait
  return(data)
})
lmm_combined <- bind_rows(lmm_data)
write_csv(lmm_combined, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/lmm_genes.csv")
```

## Big Master Sheet...

```{r}
bslmm_beak_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_beak/", pattern = "_pip01.dsv", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
bslmm_beak <- lapply(bslmm_beak_files, function(file) {
  trait <- gsub("_pip01\\.dsv", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(bslmm_beak)
bslmm_beak_all <- bind_rows(bslmm_beak)
bslmm_beak_all$model <- "bslmm"
bslmm_beak_tidy <- bslmm_beak_all[,c(9,8,2)]
View(bslmm_beak_tidy)

bslmm_feather_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_feather/", pattern = "_pip01.dsv", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
bslmm_feather <- lapply(bslmm_feather_files, function(file) {
  trait <- gsub("_pip01\\.dsv", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(bslmm_feather)
bslmm_feather_all <- bind_rows(bslmm_feather)
bslmm_feather_all$model <- "bslmm"
bslmm_feather_tidy <- bslmm_feather_all[,c(9,8,2)]
View(bslmm_feather_tidy)

bslmm_all <- rbind(bslmm_beak_tidy, bslmm_feather_tidy)

write_csv(bslmm_all, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_snps.csv")
```

```{r}
lmm_beak_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_bill/", pattern = "_outlier.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
lmm_beak <- lapply(lmm_beak_files, function(file) {
  trait <- gsub("_outlier\\.txt", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(lmm_beak)
lmm_beak_all <- bind_rows(lmm_beak)
lmm_beak_all$model <- "lmm"
lmm_beak_tidy <- lmm_beak_all[,c(15,14,2)]
View(lmm_beak_tidy)

lmm_feather_files <- list.files(path="~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_feather/", pattern = "_outlier.txt", full.names = TRUE)

# Read each file, add trait column, and combine into a single data frame
lmm_feather <- lapply(lmm_feather_files, function(file) {
  trait <- gsub("_outlier\\.txt", "", basename(file))
  data <- read.table(file, fill = TRUE, header = TRUE, sep = "\t")
  data$trait <- trait
  return(data)
})
View(lmm_feather)
lmm_feather_all <- bind_rows(lmm_feather)
lmm_feather_all$model <- "lmm"
lmm_feather_tidy <- lmm_feather_all[,c(15,14,2)]
View(lmm_feather_tidy)

lmm_all <- rbind(lmm_beak_tidy, lmm_feather_tidy)

write_csv(lmm_all, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_snps.csv")
```

```{r}
lmm_all <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/lmm/lmm_snps.csv")
bslmm_all <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/bslmm/bslmm_snps.csv")

all_snps <- rbind(bslmm_all, lmm_all)

lmm_genes <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/lmm_genes.csv")
bslmm_genes <- read.csv("~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/genes/bslmm_genes.csv")

all_genes <- rbind(bslmm_genes, lmm_genes)

master_snps <- merge(all_snps, all_genes[, c("rs", "gene_name", "gene_id")], by = "rs", all.x = TRUE)
write.csv(master_snps, "~/Desktop/GCRF/GCRF-Final/analysis/analysis_output/results_master.csv", row.names=F)
```
